Unshuffling Permutations

### 2015-09-16 article.tex
reduction

This paper is intended to study a natural generalization of the shuffle
to permutations. Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is
said to be a shuffle of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is
a shuffle of two words that are order-isomorphic to $\mu$ and $\nu$. For
example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
itself. For example, $\pi = 241356$ is a square as it is the shuffle of
$213$ with itself (\emph{e.g.},
$\pi = \dot{2}\ddot{4}\dot{1}\ddot{3}\ddot{5}\dot{6}$, where $\dot{i}$
and $\ddot{i}$), whereas $\pi = 4123$ is not.
The shuffle product defined and considered in this paper is different
from other existing and well-known shuffle products on permutations.
For instance, in~\cite{DHT:IJAC:2002}, the authors defined the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, give a permutation $\pi$, the
decision problem to test if $\pi$ is a square for the shifted shuffle
product is linear.
\smallskip

@@

Let $\pi$ be a permutation.
For the sake of clarity, we will say that a bunch of consecutive positions $P$
of $\pi$ is \emph{above} (resp. \emph{below}) above another bunch of consecutive
positions $P'$ in permutation $\pi$
if $\pi(i) > \pi(j))$ (resp. $\pi(i) < \pi(j)$) for every
$i \in P$ and every $j \in P'$.
For exemple, $\sigma_1$ is above $\sigma_2$
(in an equivalent manner, $\sigma_2$ is below $\sigma_1$)
in Fig.~\ref{subfig:increasing before and aboce decreasing},
whereas $\sigma_1$ is below $\sigma_2$
(in an equivalent manner, $\sigma_2$ is above $\sigma_1$)
in Fig.~\ref{subfig:decreasing before and below increasing}.

@@

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.45,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.45,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Lemma~\ref{lemma:at most one edge monotone}:
      Assuming a $(\nu_1,\nu_2)$-edge $(a, b)$ in $\mathcal{M}$.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

\begin{proposition}
  \label{proposition:hardness}
  Deciding whether a permutation is the union of two
  order-isomorphic patterns is \NPC.
\end{proposition}

@@

      scale=.95,

@@

    \caption{\label{fig:reduction}}

@@

  Suppose first that $\sigma$ occurs in $\pi$.
  We show how to construct a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here the fact that $\sigma$ occurs as a pattern in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in Figure~\ref{}.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here again the fact that $\sigma$ occurs as a pattern in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{}) that
  $\mathcal{M}$ is perfect and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

@@

            scale=0.45,

@@

            rectangle callout,callout relative pointer={(0.0cm,-1.5cm)}

@@

            rectangle callout,callout relative pointer={(0.0cm,3cm)}
          ] at (9.5,0)

@@

            rectangle callout,callout relative pointer={(0.0cm,-4cm)}

@@

            rectangle callout,callout relative pointer={(0.0cm,1.25cm)}

@@

        $\sigma = 312$ and $\pi = 4\mathbf{5}\mathbf{2}1\mathbf{3}6$

@@

  More precisely,
  (i) the first integer of $\sigma'$
  (\emph{i.e.}, $(k+1) + (2N_2 + N_4 + 2n + k + 2)$) is linked
  to the first integer of $\pi'$
  ((\emph{i.e.}, $(n+1) + (2N_2 + N_4 + n + k + 2)$)),
  (ii)

  Conversely, suppose that there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  We show that $\sigma$ occurs as a pattern in $\pi$.
  Whereas the matching $\mathcal{M}$ may not be as regular as in the forward
  direction, the main idea is to prove that $\mathcal{M}$ contains enough structure
  (more precisely, enough $(\sigma', \pi')$-edges) so that we can conclude.
  We have divided the reverse direction into a serie of basic claims that
  progressively defines the overall structure of $\mathcal{M}$.

>>>>

This paper is intended to study a natural generalization of the shuffle
to permutations. Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is
said to be a shuffle of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is
a shuffle of two words that are order-isomorphic to $\mu$ and $\nu$. For
example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
itself. For example, $\pi = 241356$ is a square as it is the shuffle of
$213$ with itself (\emph{e.g.},
$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
denote the second occurrence), whereas $\pi = 4123$ is not.
The shuffle product defined and considered in this paper is different
from other existing and well-known shuffle products on permutations.
For instance, in~\cite{DHT:IJAC:2002}, the authors defined the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, give a permutation $\pi$, the
decision problem to test if $\pi$ is a square for the shifted shuffle
product is linear.
\smallskip

@@

Let $\pi$ be a permutation.
For the sake of clarity, we will say that a bunch of consecutive positions $P$
of $\pi$ is \emph{above} (resp. \emph{below}) above another bunch of consecutive
positions $P'$ in permutation $\pi$
if $\pi(i) > \pi(j)$ (resp. $\pi(i) < \pi(j)$) for every
$i \in P$ and every $j \in P'$.
For exemple, $\sigma_1$ is above $\sigma_2$
(in an equivalent manner, $\sigma_2$ is below $\sigma_1$)
in Fig.~\ref{subfig:increasing before and aboce decreasing},
whereas $\sigma_1$ is below $\sigma_2$
(in an equivalent manner, $\sigma_2$ is above $\sigma_1$)
in Fig.~\ref{subfig:decreasing before and below increasing}.

@@

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.325,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad\qquad\qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.325,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Lemma~\ref{lemma:at most one edge monotone}:
      Assuming a $(\nu_1,\nu_2)$-edge $(a, b)$ in $\mathcal{M}$.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

\begin{proposition}
  \label{proposition:hardness}
  Deciding whether a permutation is a square is \NPC.
\end{proposition}

@@

      scale=.9,

@@

    \caption{\label{fig:reduction}%
      Schematic representation of the permutation $\mu$ used
      in Proposition~\ref{proposition:hardness}.
      Black edges denote the presence of at least one edge between two bunches of
      positions in $\mu$.
      Grey edges denote edges that are  only considered in the forward direction of the
      proof.
    }%

@@

  Suppose first that $\sigma$ occurs in $\pi$ and fix an occurrence.
  Construct a matching $\mathcal{M}$ as follows:
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    (i) the first position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + 1$) is linked
    to the first position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + 1$),
    (ii) the last position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + k+2$) is linked
    to the last position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + n+2$),
    and all other positions in $\sigma'$ are linked by means of $k$ pairwise
    crossing edges to the positions in
    $\pi'$ that correspond to the fixed occurence of $\sigma$ in $\pi$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    all positions in $\pi'$ that do not correspond to the fixed occurence of
    $\sigma$ in $\pi$ are linked by means of $n-k$ pairwise crossing edges
    to the positions in $\pi''$ that do not correspond to the fixed
    occurrence of $\sigma$ in $\pi$.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely, the positions in $\pi''$ that correspond to
    the fixed occurrence of $\sigma$ in $\pi$ are llinked
    linked by means of $k$ pairwise crossing edges to all positions in
    $\sigma''$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{fig:reduction}) that
  $\mathcal{M}$ is perfect, containment-free and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

@@

            scale=0.35,

@@

            rectangle callout,callout relative pointer={(0.0cm,-1cm)}

@@

            rectangle callout,callout relative pointer={(0.0cm,3.5cm)}
          ] at (9.5,-3.5)

@@

            rectangle callout,callout relative pointer={(0.0cm,-3cm)}

@@

            rectangle callout,callout relative pointer={(0.0cm,0.75cm)}

@@

        Illustration of the containment-free perfect matching $\mathcal{M}$
        between gadgets $\sigma'$, $\pi'$, $\pi''$ and $\sigma''$ assuming two
        input permutation $\sigma = 312$ and $\pi = 4\mathbf{5}\mathbf{2}1\mathbf{3}6$

@@

  Conversely, suppose that there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  We show that $\sigma$ occurs as a pattern in $\pi$.
  Whereas the matching $\mathcal{M}$ may not be as regular as in the forward
  direction, the main idea is to prove that $\mathcal{M}$ contains enough structure
  (more precisely, $k+2$ $(\sigma', \pi')$-edges) so that we can conclude
  that $\sigma$ occurs in $\pi$.
  We have divided the reverse direction into a serie of basic claims that
  progressively defines the overall structure of $\mathcal{M}$.
